GPU Efficient Texture Based Bezier Curve Evaluation

### 2015-07-18 Paper/paper.tex
I read through the paper and made some changes

\begin{abstract}
\small
Modern graphics techniques expose internal parameters to allow re-use and to help separate the technical implementation from artistic usage cases.  A popular choice is to expose parameters as customizable curves and to quantize the curves into textures.  Quantization leads to either lower quality results, or more texture memory being used to accommodate higher sampling frequencies.  The technique presented in this paper leverages the capabilities of GPU texture samplers to allow more efficient storage and evaluation of both integral and rational Bezier curves of any order, resulting in higher fidelity for the same costs.  Piecewise curves, B-Splines and NURBS are addressed, and there are also limited applications towards vector graphics.
\end{abstract}

@@

Baking curve points into a texture means that the shader program doesn't need to know what type of curve it was that made the data, nor how many control points it had, but comes at the loss of quality since there are a finite number of curve points sampled, and the data between points is just a linear interpolation between the samples.  If higher quality is desired, you can increase the resolution of the texture to be able to trade accuracy for texture memory.

Passing control point data to a shader program as shader constants allows you to get high quality curve calculations, but comes at the cost of the shader program being written in a very specific way for the type of curve you want to use, as well as more shader program instructions to calculate specific curve points.

\begin{figure}
  \includegraphics[width=5in]{Figure2.png}
  \caption{\textit{Left:} A quadratic Bezier curve written to the green channel, and the baked out equivalent using 4 pixels written to the red channel.  \textit{Right:} The technique from this paper, also using 4 pixels, written to the green channel, and the baked out equivalent written to the red channel again for comparison.}
  \label{fig:quickcomparison}
\end{figure}

@@

A quick comparison of the visual quality of these three techniques can be seen in \autoref{fig:quickcomparison}.  Besides being useful for exposing internal parameters, the technique presented in this paper is also useful as an alternative to traditional shader look up tables which use linear interpolation to get points between data points.

@@

@@

Two dimensional linear texture sampling (bilinear texture sampling) can be thought of as just the linear interpolation across the $y$ axis, of two linear interpolations across the $x$ axis.  In other words, it is just the linear interpolation of two linear interpolations.

The De Casteljeau algorithm for evaluating a quadratic curve (order 2) is also just to linear interpolate between two linear curves (order 1).  If the control points of the quadratic curve are $P_0,P_1,P_2$, let's call the first interpolation $\overline{P_0P_1}$ which is the interpolation between $P_0$ and $P_1$ for time $t$. The second linear interpolation is $\overline{P_1P_2}$ and is the interpolation between $P_1$ and $P_2$ for time $t$.  The final point on the curve will be the interpolation at time $t$ between $\overline{P_0P_1}$ and $\overline{P_1P_2}$, which we can call $\stackrel{\frown}{P_0P_1P_2}$.

We could set up a texture such that the two $x$ axis interpolations would give us the values $\overline{P_0P_1}$ and $\overline{P_1P_2}$ respectively, and then the $y$ axis interpolation between those values would give us $\stackrel{\frown}{P_0P_1P_2}$.  When sampling, we would also use texture coordinates such that the weighting is the same for the interpolations across the $x$ and $y$ axis, so that it results in the same time $t$ value for interpolation in both linear interpolations.

@@

// This function can be equivelant to the function above
// by setting the parameters:
// u = t, v = t
// A = P0, B = P1, C = P1, D = P2
float BilinearInterpolate (
  in float u, in float v,
  in float A, in float B, in float C, in float D
) {
    return mix(mix(A,B,u), mix(C,D,u), v);
}

@@

Linear texture sampling lets you control each value (each pixel), and the weighting of each level (dimension) of the interpolations, but you can set up the pixels and texture coordinates such that it evaluates a Bezier curve.  The De Casteljeau algorithm is a subset of the $N$-dimensional linear texture sampling available on the GPU.

@@

The common pixel formats contain four color channels – Red, Green, Blue and Alpha – which allows us to be able to evaluate four curves with a single texture read (\autoref{fig:texbilcurve}).

@@

@@

To see how that equation can be the same as bilinear interpolation, let's start with the equation for bilinear interpolation, which is just a linear interpolation between two other linear interpolations.  We'll use $t$ and $u$ as the interpolation weights of each interpolation.

@@

In our usage case of bilinear interpolation to evaluate Bernstein polynomials, $P_0$ will be $A$, $P_1$ and $P_2$ will both be $B$, and $P_3$ will be $C$.  Also, $t$ and $u$ are the same in our usage case so we'll just replace $u$ with $t$.  We will also replace $(1-t)$ with $s$ for readability.

@@

If we replace $A$,$B$,$C$ with $P_0$, $P_1$, $P_2$, and $s$ with $(1-t)$ we then get the familiar equation back out.

@@

This pattern continues for trilinear interpolation and beyond.  $N$ dimensional linear texture sampling is just the linear interpolation between two linear texture samples of dimension $N-1$.  Bezier curves of order $N$ are just the linear interpolation between two bezier curves of order $N-1$.

@@

Knowing that a quadratic curve is just the linear interpolation of time $t$ between the points on two linear curves at time $t$, we know that our two dimensional texture needs to set up to be able to make this happen.  The result is that if we have a $2$x$2$ texture, the top two pixels should be $P_0$ and $P_1$ and the bottom two pixels should be $P_1$ and $P_2$.  When we do a bilinear interpolation such that we plug $t$ into both $u$ and $v$, the end result will be that $\overline{P_0P_1}$ and $\overline{P_1P_2}$ will be calculated from the $u$ axis interpolation, and then those two will be linearly interpolated across the $v$ axis to give us the final result.

@@

@@

Knowing that a cubic curve is just the linear interpolation of time $t$ between the points on two quadratic curves at time $t$, we know that our three dimensional texture needs to be set up to be able to make this happen.  With a $2$x$2$x$2$ texture, the front $2$x$2$ pixels would encode a quadratic curve for $P_0,P_1,P_2$ and the back $2$x$2$ pixels would encode a quadratic curve for $P_1,P_2,P_3$.  Then, when we sample the texture, plugging $t$ into $u,v$ and $w$, the result will be that it will calculate the points on the quadratic curves $P_0,P_1,P_2$ at time $t$ and also $P_1,P_2,P_3$ at time $t$, and then will interpolate across the $w$ axis to give us the final result.

@@

@@

Rational Bezier curves are useful because they can be used to make shapes which cannot be formed from integral Bezier curves, including perfectly representing conic sections which mean that they can also represent sine and cosine as seen in \autoref{fig:rationalbeziersincos}.  Rational Bezier curves are defined by the equation:

@@

Looking closely, we can see that we essentially have one Bezier curve divided by another.  The numerator has a weighting and a control point, while the denominator has only a weighting, but if we were to multiply the weighting into the control point in the numerator, we would essentially have one Bezier curve divided by another.

Using this technique, we can easily encode the numerator into one color channel, and the denominator into another.

@@

So far we have only talked about one dimensional - or explicit - Bezier curves which take the form of $y=f(x)$.  Multidimensional Bezier curves are evaluated separately per axis though, which means that you can encode information per axis in different color channels.

@@

For instance, if you had a 2x2 image which encoded a quadratic Bezier curve, you might do two texture reads, letting the texture sampler do the $x$ axis interpolation, and then you could linearly interpolate those results in the shader program in full precision floating point math.  This would give you a result somewhere between full hardware and full software interpolation.  You can compare the results in \autoref{fig:samplehwsw}.

@@

Explicit Bezier curves are curves which have scalar control points and are in the form of $y=f(x)$.  For these curves, when you plug in a value for $x$ you get a $y$ value out.  They are unable to make curves that have multiple $y$ values for the same $x$, and the curves can't cross over themselves to make closed shapes.  This does make them not as useful to vector graphics as 2D Bezier curves which are of the form $(x,y) = f(t)$, but they still do have some use cases.  Rendering explicit bezier curves using this technique is very simple.  If rendering a quad, you just choose either the $x$ or $y$ axis, normalize it from 0 to 1 and use that to calculate the other axis value by doing a texture sample.  At this point, you can either choose a half space to color in, or you can calculate the gradient of the function (by using the built in dFdx style functions to get partial derivatives), and use that as an estimated distance to the curve.

>>>>

\begin{abstract}
\small
Modern graphics techniques expose internal parameters to allow re-use and to help separate the technical implementation from artistic usage cases.  A popular choice is to expose parameters as customizable curves and to quantize the curves into textures.  Quantization leads to either lower quality results, or more texture memory being used to accommodate higher sampling frequencies.  The technique presented in this paper leverages the capabilities of GPU texture samplers to allow more efficient storage and evaluation of both integral and rational Bezier curves of any order, resulting in higher fidelity for similar costs.  Piecewise curves, B-Splines and NURBS are addressed, and there are also limited applications towards vector graphics.
\end{abstract}

@@

Baking curve points into a texture means that the shader program doesn't need to know what type of curve it was that made the data, nor how many control points it had, but comes at the loss of quality since there are a finite number of curve points sampled, and the data between points is just a linear interpolation between the samples.  If higher quality is desired, you can increase the resolution of the texture to trade accuracy for texture memory.

Passing control point data to a shader program as shader constants allows you to get high quality curve calculations, but comes at the cost of the shader program being written specifically for the type of curve you want to use, and also takes more shader program instructions to calculate specific curve points.

\begin{figure}
  \includegraphics[width=5in]{Figure2.png}
  \caption{Quadratic Bezier curves (yellow) compared to a 4x1 baked out (quantized) texture for the same curve (red).  \textit{Left:} Curve evaluated from shader program constants. \textit{Right:}  The technique from this paper - the curve is stored in a 2x2 texture and a bilinear texture sample is taken to get each point on the curve.}
  \label{fig:quickcomparison}
\end{figure}

@@

A quick comparison of the visual quality of these three techniques can be seen in \autoref{fig:quickcomparison}.

Besides being useful for exposing internal parameters, the technique presented in this paper is also useful as an alternative to traditional shader look up tables, since this technique allows non linear interpolation between data points, where traditional lookup table textures only offer linear interpolation between data points.

@@

TODO: clean up the notation here!

@@

Two dimensional linear texture sampling (bilinear texture sampling) can be thought of as just the linear interpolation across the $y$ axis, of two linear interpolations across the $x$ axis (or an interpolation across the $x$ axis of two linear interpolations across the $y$ axis - order doesn't matter).  In other words, bilinear texture sampling is just the linear interpolation of two linear interpolations.

The De Casteljeau algorithm for evaluating a quadratic curve (order 2) is also just a linear interpolation between two linear interpolations.  If the control points of the quadratic curve are $P_0,P_1,P_2$, let's call the first interpolation $\overline{P_0P_1}$ which is the interpolation between $P_0$ and $P_1$ for time $t$. The second linear interpolation is $\overline{P_1P_2}$ and is the interpolation between $P_1$ and $P_2$ for time $t$.  The final point on the curve will be the interpolation at time $t$ between $\overline{P_0P_1}$ and $\overline{P_1P_2}$, which we can call $\stackrel{\frown}{P_0P_1P_2}$.

We could set up a texture such that the two $x$ axis interpolations would give us the values $\overline{P_0P_1}$ and $\overline{P_1P_2}$ respectively, and then the $y$ axis interpolation between those values would give us $\stackrel{\frown}{P_0P_1P_2}$.  When sampling, we would also use texture coordinates such it results in the same time $t$ value for interpolation on each axis.

@@

// This function is equivelant to the function above when:
// u = t, v = t
// A = P0, B = P1, C = P1, D = P2
float BilinearInterpolate (
  in float u, in float v,
  in float A, in float B, in float C, in float D
) {
    return mix(mix(A,B,u), mix(C,D,u), v);
}

@@

Compared to a quadratic curve evaluated with De Casteljeau's algorithm, bilinear texture sampling has two blend weights instead of one, and has four values to interpolate between instead of three, but you can set a texture up such that when you sample it a specific way, you get as output a quadratic Bezier curve.  Because of this, the De Casteljeau algorithm is a subset of the $N$-dimensional linear texture sampling available on the GPU.

@@

The common pixel formats contain four color channels –-- Red, Green, Blue and Alpha –-- which allows us to be able to evaluate four curves with a single texture read (\autoref{fig:texbilcurve}).

@@

TODO: fix up notation to be consistant (Px vs A,B,C,D)

@@

To see how that equation can be the same as bilinear interpolation, let's start with the equation for bilinear interpolation, which is just a linear interpolation between two other linear interpolations.  We'll use $t$ and $u$ as the interpolation weights of each interpolation.  We will linearly interpolate between linear interpolations $E(u)$ and $F(u)$ by $t$ to get out a result of $P(t,u)$

@@

In our usage case of bilinear interpolation to evaluate Bernstein polynomials, $P_0$ will be $A$, $P_1$ and $P_2$ will both be $B$, and $P_3$ will be $C$.  Also, we set $t$ and $u$ equal in our usage case so we'll replace $u$ with $t$.  We will also replace $(1-t)$ with $s$ for readability.

@@

If we replace $A$,$B$,$C$ with $P_0$, $P_1$, $P_2$, and $s$ with $(1-t)$ we then get the familiar equation below, which is the Bernstein form of a quadratic Bezier curve.

@@

This pattern continues for trilinear interpolation mapping to a cubic curve, and beyond.  $N$ dimensional linear texture sampling is just the linear interpolation between two linear texture samples of dimension $N-1$.  Bezier curves of order $N$ are just the linear interpolation between two bezier curves of order $N-1$.

@@

Knowing that a quadratic curve is just the linear interpolation of time $t$ between the points on two linear curves at time $t$, we have to find a way to make a texture that will cause this to happen with bilinear interpolation.  The result is that if we have a $2$x$2$ texture, the top two pixels should be $P_0$ and $P_1$ and the bottom two pixels should be $P_1$ and $P_2$.  When we do a bilinear interpolation such that we plug $t$ into both $u$ and $v$, the end result will be that $\overline{P_0P_1}$ and $\overline{P_1P_2}$ will be calculated from the $u$ axis interpolation, and then those two will be linearly interpolated across the $v$ axis to give us the final result.

@@

GLSL source code is also provided below to help see how that would work for sampling a cubic curve from a two dimensional texture, with two bilinear texture samples combined to create the final cubic curve.

@@

Knowing that a cubic curve is just the linear interpolation of time $t$ between the points on two quadratic curves at time $t$, we need to find a way to set up a 3d texture to make this happen with trilinear texture sampling.  With a $2$x$2$x$2$ texture, the front $2$x$2$ pixels would encode a quadratic curve for $P_0,P_1,P_2$ and the back $2$x$2$ pixels would encode a quadratic curve for $P_1,P_2,P_3$.  Then, when we sample the texture, plugging $t$ into $u,v$ and $w$, the result will be that it will calculate the points on the quadratic curves $P_0,P_1,P_2$ at time $t$ and also $P_1,P_2,P_3$ at time $t$, and then will interpolate across the $w$ axis to give us the final result.

@@

Below is some GLSL source code showing how you would sample a cubic curve stored in a 3d 2x2x2 texture, using trilinear texture sampling.

@@

Rational Bezier curves are useful because they can be used to make shapes which cannot be formed from integral Bezier curves, including perfectly representing conic sections.  This also means that they can accurately represent sine and cosine as seen in \autoref{fig:rationalbeziersincos}.  Rational Bezier curves are defined by the equation:

@@

If we were to multiply the weight into the control point for the numerator, We can see that we essentially have one Bezier curve divided by another.  Using the techniques in this paper, we could easily encode the numerator into one color channel, and the denominator into another.

@@

So far we have only talked about one dimensional --- or explicit --- Bezier curves which take the form of $y=f(x)$.  Multidimensional Bezier curves are evaluated separately per axis though, which means that you can encode information per axis in different color channels.

@@

For instance, if you had a 2x2 image which encoded a quadratic Bezier curve, you might do two texture reads, letting the texture sampler do the $x$ axis interpolation, and then you could linearly interpolate those results in the shader program in full precision floating point math for the $y$ axis.  This would give you a result somewhere between full hardware and full software interpolation.  You can compare the results in \autoref{fig:samplehwsw}.

@@

Explicit Bezier curves are curves which have scalar control points and are in the form of $y=f(x)$.  For these curves, when you plug in a value for $x$ you get a $y$ value out.  They are unable to make curves that have multiple $y$ values for the same $x$, and the curves can't cross over themselves to make closed shapes.  This does make them not as useful to vector graphics as 2D Bezier curves which are of the form $(x,y) = f(t)$, but they still do have some use cases.  Rendering explicit Bezier curves using this technique is very simple.  If rendering a quad, you just choose either the $x$ or $y$ axis, normalize it from 0 to 1 and use that to calculate the other axis value by doing a texture sample.  At this point, you can either choose a half space to color in, or you can calculate the gradient of the function (by using the built in dFdx style functions to get partial derivatives), and use that as an estimated distance to the curve.
